How to build an homogeneous language for heterogeneous platforms

### 2013-06-05 papers/gpce2013/gpce2013.tex
Better abstract and intro. WIP contrib.

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between client and server code. Our DSLs have a
performance / expressiveness ratio about two times higher than other approaches.
\end{abstract}

@@

\keywords Heterogeneous Code Generation, Domain-specific languages

@@

The more interactive the application is, the more logic needs to be duplicated between the server-side
and the client-side, and the higher is the amount of client-side code. Developers use libraries and frameworks to get
high-level abstractions on client-side, making their code easier to reason about and to maintain, but also making
their code run less efficiently (abstraction penalty).

@@

Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the translation of
common abstractions like collections into their native counterpart (JavaScript arrays on client-side and standard
library's collections on server-side) may be hard coded in the compiler, but that would not scale to handle all the
abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even some
business data type that may be represented differently for performance reasons).

On one hand, for engineering reasons, developers want to write Web applications using a single high-level language,
abstracting over the target platforms differences. But on the other hand, for performance reasons, they want to keep
control on the way their code is compiled to each target platform. How to solve this dilemma?

Compiled domain specific embedded languages~\cite{Elliott2003_Compiling} allow the definition of domain specific
languages (DSLs) as libraries on top of a host language, and to compile them to a target platform. The deep embedding
gives the opportunity to control the code generation scheme for a given abstraction and target platform.

\texttt{js-scala} is such a compiled embedded DSL defined in Scala that generates JavaScript code, making it possible
to write the client-side code of Web applications using JavaScript~\cite{Kossakowski12_JsDESL}. This paper
enriches \texttt{js-scala} in order to solve the problem described above. We use staging~\cite{Jorring1986_Staging}
to:

\begin{itemize}
 \item generate efficient code for typical abstractions used in Web programming;
 \item generate specialized code for both client and server sides for shared abstractions.
\end{itemize}

Though the code written in \texttt{js-scala} is high-level and can be shared between clients and
servers, it has the same runtime performances as hand-tune low-level JavaScript code. We observed a performance /
expressiveness ratio two times higher than using other approaches.

@@

Another approach consists in defining several independent domain-specific languages~\cite{Van00_DSL}, each one
focusing on concerns specific to a given problem domain, and then to combine all the source artifacts written with
these language into one executable program, as shown in figure~\ref{langs} (b). Defining such a language requires a
minimal effort compared to the previous approach. On the other hand, it is difficult to have interoperability between
DSLs (ref). \cite{Groenewegen08_WebDSL} shows such a domain-specific lanugage for defining Web applications.

@@

Lightweight Modular Staging~\cite{Rompf12_LMSThesis} is a framework for defining deeply embedded DSLs in Scala. It
has been used to define high-performance DSLs for parallel computing~\cite{Brown11_Parallel} and can be used to
generate JavaScript code~\cite{Kossakowski12_JsDESL}.

\section{High-Level Abstractions for Client-Side Code}
\label{contribution}

In a Web application, the user interface is defined by a HTML document that can be updated by the JavaScript code.
A typical operation consists in searching some “interesting” element in the document, in order to extract its
content, replace it or listen to user events triggered on it. The standard API provides several functions to search
elements in a HTML document according to their name or attribute values. Figure~\ref{selectors-api} summarizes the
available functions and their differences.

@@

The \code{querySelector} and \code{querySelectorAll} are the most general functions, the others handle special
cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
most JavaScript developers use the jQuery
library~\cite{Bibeault08_jQuery}\footnote{According to the following document:
\href{http://trends.builtwith.com/javascript}{ http://trends.builtwith.com/javascript}, jQuery is used by more than
40\% of the top million sites} that provides only one function to search for elements.
Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs, the first one
using the native APIs and the second one using jQuery.

@@

Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using js-scala. We provide two
functions for searching elements: \code{find} to find the first element matching a selector and \code{findAll} to
find all the matching elements. During staging these functions analyze the selector that is passed as parameter and,
when possible, produce code using the specialized API, otherwise they produce code using \code{querySelector} and
\code{querySelectorAll}. Thanks to this optimizations, listing~\ref{js-scala-selectors} generates a JavaScript
program identical to listing~\ref{vanilla-selectors}. Writing the program with js-scala instead of using a JavaScript
library such as jQuery is interesting because the abstraction for searching elements exist only in the initial
program source code, not in the final JavaScript program. It gives two advantages: (1) the execution of the final
program is more performant because of the use of specialized APIs, (2) the final program’s size is smaller because it
does not need to include jQuery.

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
button in the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
selector in the document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Listing \ref{null-defensive} shows defensive code to handle \code{null}
cases, but it leads to very cumbersome code.

@@

@@

>>>>

\begin{abstract}
Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
and abstracting over the clients and servers differences without trading this engineering comfort for performance?
This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
code leveraging the target platform characteristics, (2) be shared between client and server code. Code written with
our DSL has a high level of abstraction but runs as fast as hand tuned low-level JavaScript code.
\end{abstract}

@@

\keywords Heterogeneous code generation, Domain-specific languages

@@

The more interactive the application is, the more logic needs to be duplicated between the server-side
and the client-side, and the higher is the complexity of the client-side code. Developers use libraries and
frameworks to get high-level abstractions on client-side, making their code easier to reason about and to maintain,
but also making their code run less efficiently (due to \emph{abstraction penalty}).

@@

Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the translation of
common abstractions like collections into their native counterpart (JavaScript arrays on client-side and standard
library's collections on server-side) may be hard coded in the compiler, but that would not scale to handle all the
abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even some
business data type that may be represented differently).

On one hand, for engineering reasons, developers want to write Web applications using a single high-level language,
abstracting over the target platforms differences and reducing code complexity. But on the other hand, for
performance reasons, they want to keep control on the way their code is compiled to each target platform. How to
solve this dilemma?

Compiled domain-specific embedded languages~\cite{Elliott2003_Compiling} allow the definition of domain-specific
languages (DSLs) as libraries on top of a host language, and to compile them to a target platform. The deep embedding
gives the opportunity to control the code generation scheme for a given abstraction and target platform.

Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that generates JavaScript
code, making it possible to write the client-side code of Web applications using
JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the engineering dilemma described
above. This paper enriches \emph{js-scala} with the following contributions:

\begin{itemize}
 \item we define high-level abstractions typically used in Web programming generating low-level code leveraging
native Web browsers APIs;
 \item in the case of abstractions shared between servers and clients, we specialize the code generation in order to
leverage the target platform native environments.
\end{itemize}

We validate our approach with a case study implemented with various candidate technologies and discuss the relative
pro and cons of them. Though the code written in our DSL is high-level and can be shared between clients and servers,
it has the same runtime performances as hand-tune low-level JavaScript code.

@@

Another approach consists in defining several independent domain-specific languages~\cite{Van00_DSL}, each one
focusing on concerns specific to a given problem domain, and then to combine all the source artifacts written with
these language into one executable program, as shown in figure~\ref{langs} (b). Defining such a language requires a
minimal effort compared to the previous approach. On the other hand, it is difficult to have interoperability between
DSLs (ref). \cite{Groenewegen08_WebDSL} shows such a domain-specific language for defining Web applications.

@@

Lightweight Modular Staging~\cite{Rompf12_LMSThesis} is a framework for defining deeply embedded DSLs in Scala using
staging~\cite{Jorring1986_Staging}. It has been used to define high-performance DSLs for parallel
computing~\cite{Brown11_Parallel}.

\section{High-Level Abstraction for Client-Side Code: Selectors}
\label{contribution}

In a Web application, the user interface is defined by a HTML document that can be updated by the JavaScript code.
A typical operation consists in searching some “interesting” element in the document, in order to extract its
content, replace it or listen to user events triggered on it (such as a mouse click). The standard API provides
several functions to search elements in a HTML document according to their name or attribute values.
Figure~\ref{selectors-api} summarizes the available functions and their differences.

@@

The \code{querySelector} and \code{querySelectorAll} are the most general functions while the others handle special
cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
most JavaScript developers use the jQuery
library~\cite{Bibeault08_jQuery}\footnote{According to the following document:
\href{http://trends.builtwith.com/javascript}{ http://trends.builtwith.com/javascript}, jQuery is used by more than
40\% of the top million sites} that provides only one high-level function to search for elements.
Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs performing
element searches, the first one using the native APIs and the second one using jQuery.

@@

Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using \emph{js-scala}. We
provide two functions for searching elements: \code{find} to find the first element matching a selector and
\code{findAll} to find all the matching elements. During staging these functions try to analyze the selector that is
passed as parameter and, when appropriate, produce code using the specialized API, otherwise they produce code using
\code{querySelector} and \code{querySelectorAll}. As a result, listing~\ref{js-scala-selectors} generates a
JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level abstraction (the \code{find} and
\code{findAll} functions) exist only in the initial program, not in the final JavaScript program. It gives two
advantages: (1) the execution of the final program is more performant because of the use of specialized APIs, (2) the
final program’s size is smaller because it does not need to include jQuery. (Move this last sentence in the
discussion)

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
button in the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
selector in the document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Defensive code can handle \code{null} reference, but leads to very cumbersome
code, as shown in listing \ref{null-defensive}.

@@

Some programming languages encode optional values with a monad (\eg \code{Maybe} in Haskell and \code{Option} in
Scala). In that case, sequencing over the monad encodes optional value dereferencing. If the language supports a
convenient syntax for monad sequencing, it brings a convenient syntax for optional value dereferencing, alleviating
developers from the burden of defensive programming.

Listing \ref{null-js-scala} implements in \emph{js-scala} a program equivalent to listing \ref{null-defensive}. The
\code{for} notation is used to sequence computations over optional values encoded with a monad, allowing a high
level of expressiveness. (To be continued, mention \code{Rep[Option[A]]}, \code{Rep[Some[A]]}, \code{Rep[None.type]})

@@

\subsection{DOM Fragments Definition}
